Data Specialization for a Staged Lambda Calculus

### 2015-07-03 paper/semantics.tex
fixed up semantics section

Differing worlds (and hence, differing stages of computation) interact by means
of the $\fut$ and $\curr$ type formers.  These modalities are internalizations
of typing at worlds \bbtwo\ and \bbonep, respectively, to typing at \bbonep.
For example, given a term $e$ of type $A$ at world \bbtwo, 
$\next{e}$ has type $\fut A$ at \bbonem. This essentially
encapsulates $e$ as a computation to be evaluated in the future---first-stage
computations can shuffle it around, but not use its result.  The only way to use
a $\fut A$ is to wrap it with $\prev$, yielding an $A$ at \bbtwo.

@@

The central tenet of a staged language
is that first stage code should be evaluated entirely before second stage code.
Accordingly, our dynamics operates in two passes.
The first pass takes an input program $\coltwo e A$ and reduces it to a residual,
$\exv q$, and the second pass further reduces this residual.
Since a residual $q$ is a monostage objects, this second pass is just standard monostage evaluation.
Consequently, this paper is concerned only with the first pass of evaluation.

@@

\paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgement $\steptwo e {e'}$.
Since no reduction of stage two terms is to take place, this judgement does nothing but search for \prev\ blocks within $e$.
A world \bbtwo\ term is done evaluating when it has the form $\exv q$, where $q$ is a residual. 
To be a residual, a term must have no first stage subterms (equivalently, no $\prev$s), even within the body of a function or branches of a case.
This implies that $\stepsym \bbtwo$ must proceed underneath second stage binders.

@@

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, searches for redexes and reduces them in place.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other kinds of steps there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated when partial values are substituted for variable,
as they would be in the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
For example, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $\hat{y}$, and substituting that variable instead:
\begin{lstlisting} 
2`let val yhat = e in prev{`1`[next{`2`yhat`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgement, called a hoisting step.
These steps are introduced when a \next\ block has contents that are a residual but not already a variable.
In essence, they operate by ``bubbling up'' a subtitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.

@@

@@

The multistage function \texttt{qss} from \ref{sec:overview} has
type $A\to \fut (B \to C)$ at \bbonem.%
\footnote{We can rewrite \texttt{qss} in this curried form, or apply a
higher-order currying function.}
To apply this function to arguments $\colmix a A$ and $\coltwo b B$, simply
evaluate the program:
\begin{lstlisting}
2`prev{`1`f a`2`} b`
\end{lstlisting}

We can also directly observe the reuse of first-stage computations across
multiple second-stage computations. To run many order statistics queries
$\texttt{k1},\dots,\texttt{km}$ on the same list, run the program:
\begin{lstlisting}
2`let val qs = prev{`1`qss `3`gr{`1`[7,4,2,5,9,...,3]`3`}`2`}`
2`in (qs k1,...,qs km)`
\end{lstlisting}
To see that the partial application of \texttt{qss} is only evaluated once,
observe that after the first stage of evaluation, the residual has the form:
\begin{lstlisting}
2`let val yhat = ...`
2`    val qs = yhat`
2`in (qs k1,...,qs km)`
\end{lstlisting}
where $\hat{y}$ is bound to a function containing no first-stage redexes.

>>>>

Differing worlds (and hence, differing stages of computation) interact by means
of the $\fut$ and $\curr$ type formers.  These modalities are internalizations
of typing at worlds \bbtwo\ and \bbonep, respectively, to typing at \bbonem.
For example, given a term $e$ of type $A$ at world \bbtwo, 
$\next{e}$ has type $\fut A$ at \bbonem. 
This essentially encapsulates $e$ as a computation that will be evaluated in the future, 
but it provides a handle (of type $\fut A$) to that eventual value which can be moved around now.
Computations at \bbonem\ can shuffle this handle around, but not access the future result it refers to.  
The only way to eliminate a $\fut$ wrapper is to do so from the future by using a {\tt prev}, yielding an $A$ at \bbtwo.

@@

The central tenet of a staged language
is that first stage code should be evaluated entirely before second stage code.
Accordingly, our dynamics operates in two passes.
The first pass takes an input program $\coltwo e A$
and reduces all of its first-stage subterms in place,
eventually resulting in a residual $\exv q$.
The second pass further reduces this residual.
Since a residual $q$ is a monostage object, this second pass is just standard monostage evaluation.
Consequently, this paper is concerned only with the first pass of evaluation.

@@

\paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgement \mbox{$\steptwo e {e'}$}.
Since no reduction of stage two terms is to take place, this judgement does nothing but search for \prev\ blocks within $e$.
A world \bbtwo\ term is done evaluating when it has the form $\exv q$, where $q$ is a residual. 
To be a residual, a term must have no first stage subterms (equivalently, no $\prev$s), even within the body of a function or branches of a case.
This implies that $\stepsym \bbtwo$ must proceed underneath second stage binders.

@@

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, searches for redexes and reduces them in place.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other kinds of steps there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated when partial values are substituted for variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
We find this interpretation to be more natural for our applications.
For example, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgement, called a \emph{hoisting step}.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a subtitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.

@@

\TODO pull out the theorems into a figure

@@

The multistage function \texttt{qss} from \ref{sec:overview} has
type fitting the pattern \mbox{$A\to \fut (B \to C)$} at \bbonem.%
\footnote{We can rewrite \texttt{qss} in this curried form, or apply a
higher-order currying function.} 
To apply this function to arguments $\colmix a A$ and $\coltwo b B$, simply
evaluate the program:
\begin{lstlisting}
2`prev{`1`f a`2`} b`
\end{lstlisting}

We can also directly observe the reuse of first-stage computations across
multiple second-stage computations. To run many order statistics queries
$\texttt{k1},\dots,\texttt{km}$ on the same list, run the program:
\begin{lstlisting}
2`let val qs = prev{`1`qss `3`gr{`1`[7,4,2,5,9,...,3]`3`}`2`}`
2`in (qs k1,...,qs km)`
\end{lstlisting}
To see that the partial application of \texttt{qss} is only evaluated once,
observe that after the first stage of evaluation, the residual has the form:
\begin{lstlisting}
2`let val y = ...`
2`    val qs = y`
2`in (qs k1,...,qs km)`
\end{lstlisting}
where $\hat{y}$ is bound to a function containing no first-stage redexes.
